---
title: "Data Types and Structures"
keypoints:
- R's basic data types are character, numeric, integer, complex, and logical.
- R's basic data structures include the vector, list, matrix, data frame, and factors.
- Objects may have attributes, such as name, dimension, and class.
objectives:
- Expose learners to the different data types in R.
- Learn how to create vectors of different types.
- Be able to check the type of vector.
- Learn about missing data and other special values.
- Getting familiar with the different data structures (lists, matrices, data frames).
questions:
- What are the different data types in R?
- What are the different data structures in R?
- How do I access data within the various data structures?
- How to use lubridate package to handle date and time?
teaching: 45
exercises: 0
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("13-supp-data-structures-")
```

### Understanding Basic Data Types in R

To make the best of the R language, you'll need a strong understanding of the
basic data types and data structures and how to operate on those.

Very important to understand because these are the objects you will manipulate
on a day-to-day basis in R. Dealing with object conversions is one of the most
common sources of frustration for beginners.

**Everything** in R is an object.

R has 6 (although we will not discuss the raw class for this workshop) atomic
vector types.

* character
* numeric (real or decimal)
* integer
* logical
* complex

By *atomic*, we mean the vector only holds data of a single type.

* **character**: `"a"`, `"swc"`
* **numeric**: `2`, `15.5`
* **integer**: `2L` (the `L` tells R to store this as an integer)
* **logical**: `TRUE`, `FALSE`
* **complex**: `1+4i` (complex numbers with real and imaginary parts)

R provides many functions to examine features of vectors and other objects, for
example

* `class()` - what kind of object is it (high-level)?
* `typeof()` - what is the object's data type (low-level)?
* `length()` - how long is it? What about two dimensional objects?
* `attributes()` - does it have any metadata?

```{r}
# Example
x <- "dataset"
typeof(x)
attributes(x)

y <- 1:10
y
typeof(y)
length(y)

z <- as.numeric(y)
z
typeof(z)
```

R has many __data structures__. These include

* atomic vector
* list
* matrix
* data frame
* factors

### Atomic Vectors

A vector is the most common and basic data structure in R and is pretty much the
workhorse of R. Technically, vectors can be one of two types:

* atomic vectors
* lists

although the term "vector" most commonly refers to the atomic types not to lists.

### The Different Vector Modes

A vector is a collection of elements that are most commonly of mode `character`,
`logical`, `integer` or `numeric`.

You can create an empty vector with `vector()`. (By default the mode is
`logical`. You can be more explicit as shown in the examples below.) It is more
common to use direct constructors such as `character()`, `numeric()`, etc.

```{r}
vector() # an empty 'logical' (the default) vector
vector("character", length = 5) # a vector of mode 'character' with 5 elements
character(5) # the same thing, but using the constructor directly
numeric(5)   # a numeric vector with 5 elements
logical(5)   # a logical vector with 5 elements
```

You can also create vectors by directly specifying their content. R will then
guess the appropriate mode of storage for the vector. For instance:

```{r}
x <- c(1, 2, 3)
```

will create a vector `x` of mode `numeric`. These are the most common kind, and
are treated as double precision real numbers. If you wanted to explicitly create
integers, you need to add an `L` to each element (or *coerce* to the integer
type using `as.integer()`).

```{r}
x1 <- c(1L, 2L, 3L)
```

Using `TRUE` and `FALSE` will create a vector of mode `logical`:

```{r}
y <- c(TRUE, TRUE, FALSE, FALSE)
```

While using quoted text will create a vector of mode `character`:

```{r}
z <- c("Sarah", "Tracy", "Jon")
```

### Examining Vectors

The functions `typeof()`, `length()`, `class()` and `str()` provide useful
information about your vectors and R objects in general.

```{r}
typeof(z)
length(z)
class(z)
str(z)
```

> ## Finding Commonalities
>
> Do you see a property that's common to all these vectors above?
{: .challenge}

### Adding Elements

The function `c()` (for combine) can also be used to add elements to a vector.

```{r}
z <- c(z, "Annette")
z
z <- c("Greg", z)
z
```

### Vectors from a Sequence of Numbers

You can create vectors as a sequence of numbers.

```{r}
series <- 1:10
seq(10)
seq(from = 1, to = 10, by = 0.1)
```

### Missing Data

R supports missing data in vectors. They are represented as `NA` (Not Available)
and can be used for all the vector types covered in this lesson:

```{r}
x <- c(0.5, NA, 0.7)
x <- c(TRUE, FALSE, NA)
x <- c("a", NA, "c", "d", "e")
x <- c(1+5i, 2-3i, NA)
```

The function `is.na()` indicates the elements of the vectors that represent
missing data, and the function `anyNA()` returns `TRUE` if the vector contains
any missing values:

```{r}
x <- c("a", NA, "c", "d", NA)
y <- c("a", "b", "c", "d", "e")
is.na(x)
is.na(y)
anyNA(x)
anyNA(y)
```

### Other Special Values

`Inf` is infinity. You can have either positive or negative infinity.

```{r}
1/0
```

`NaN` means Not a Number. It's an undefined value.

```{r}
0/0
```

### What Happens When You Mix Types Inside a Vector?

R will create a resulting vector with a mode that can most easily accommodate
all the elements it contains. This conversion between modes of storage is called
"coercion". When R converts the mode of storage based on its content, it is
referred to as "implicit coercion". For instance, can you guess what the
following do (without running them first)?

```{r}
xx <- c(1.7, "a")
xx <- c(TRUE, 2)
xx <- c("a", TRUE)
```

You can also control how vectors are coerced explicitly using the
`as.<class_name>()` functions:

```{r}
as.numeric("1")
as.character(1:2)
```

### Objects Attributes

Objects can have __attributes__. Attributes are part of the object. These include:

* names
* dimnames
* dim
* class
* attributes (contain metadata)

You can also glean other attribute-like information such as length (works on
vectors and lists) or number of characters (for character strings).

```{r}
length(1:10)
nchar("Software Carpentry")
```

### Matrix

In R matrices are an extension of the numeric or character vectors. They are not
a separate type of object but simply an atomic vector with dimensions; the
number of rows and columns.

```{r}
m <- matrix(nrow = 2, ncol = 2)
m
dim(m)
```

Matrices in R are filled column-wise.

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
```

Other ways to construct a matrix

```{r}
m      <- 1:10
dim(m) <- c(2, 5)
```

This takes a vector and transforms it into a matrix with 2 rows and 5 columns.

Another way is to bind columns or rows using `cbind()` and `rbind()`.

```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
```

You can also use the `byrow` argument to specify how the matrix is filled. From R's own documentation:

```{r}
mdat <- matrix(c(1, 2, 3, 11, 12, 13),
               nrow = 2,
               ncol = 3,
               byrow = TRUE)
mdat
```

Elements of a matrix can be referenced by specifying the index along each dimension (e.g. "row" and "column") in single square brackets.

```{r}
mdat[2, 3]
```

### List

In R lists act as containers. Unlike atomic vectors, the contents of a list are
not restricted to a single mode and can encompass any mixture of data
types. Lists are sometimes called generic vectors, because the elements of a
list can by of any type of R object, even lists containing further lists. This
property makes them fundamentally different from atomic vectors.

A list is a special type of vector. Each element can be a different type.

Create lists using `list()` or coerce other objects using `as.list()`. An empty
list of the required length can be created using `vector()`

```{r}
x <- list(1, "a", TRUE, 1+4i)
x

x <- vector("list", length = 5) # empty list
length(x)
```

The content of elements of a list can be retrieved by using double square brackets.

```{r}
x[[1]]
```

Vectors can be coerced to lists as follows:

```{r}
x <- 1:10
x <- as.list(x)
length(x)
```

1. What is the class of `x[1]`?
2. What about `x[[1]]`?


Elements of a list can be named (i.e. lists can have the `names` attribute)

```{r}
xlist <- list(a = "Karthik Ram", b = 1:10, data = head(iris))
xlist
names(xlist)
```

1. What is the length of this object? What about its structure?

Lists can be extremely useful inside functions. Because the functions in R are able to return only a single object, you can "staple" together lots
of different kinds of results into a single object that a function can return.

A list does not print to the console like a vector. Instead, each element of the
list starts on a new line.

Elements are indexed by double brackets. Single brackets will still return
a(nother) list. If the elements of a list are named, they can be referenced by the `$` notation (i.e. `xlist$data`).

### Dates and times in R
Lubridate is an R package that makes it easier to work with dates and times. Below is a concise tour of some of the things lubridate can do for you. Lubridate was created by Garrett Grolemund and Hadley Wickham, and is now maintained by Vitalie Spinu.

1. Parsing dates and times
Getting R to agree that your data contains the dates and times you think it does can be tricky. Lubridate simplifies that. Identify the order in which the year, month, and day appears in your dates. Now arrange “y”, “m”, and “d” in the same order. This is the name of the function in lubridate that will parse your dates. For example,

```{r}
install.packages("lubridate")
library(lubridate)
ymd("20110604")
mdy("06-04-2011")
dmy("04/06/2011")
```
Lubridate’s parse functions handle a wide variety of formats and separators, which simplifies the parsing process.

If your date includes time information, add h, m, and/or s to the name of the function. ymd_hms is probably the most common date time format. To read the dates in with a certain time zone, supply the official name of that time zone in the tz argument.

```{r}
arrive <- ymd_hms("2011-06-04 12:00:00", tz = "Pacific/Auckland")
arrive
leave <- ymd_hms("2011-08-10 14:00:00", tz = "Pacific/Auckland")
leave
```
2. Setting and Extracting information
Extract information from date times with the functions second, minute, hour, day, wday, yday, week, month, year, and tz. You can also use each of these to set (i.e, change) the given information. Notice that this will alter the date time. wday and month have an optional label argument, which replaces their numeric output with the name of the weekday or month.

```{r}
second(arrive)
second(arrive) <- 25
arrive


second(arrive) <- 0
wday(arrive)
wday(arrive, label = TRUE)
```

3. Time Zones
There are two very useful things to do with dates and time zones. First, display the same moment in a different time zone. Second, create a new moment by combining an existing clock time with a new time zone. These are accomplished by with_tz and force_tz.

For example, a while ago I was in Auckland, New Zealand. I arranged to meet the co-author of lubridate, Hadley, over skype at 9:00 in the morning Auckland time. What time was that for Hadley who was back in Houston, TX?

```{r}
meeting <- ymd_hms("2011-07-01 09:00:00", tz = "Pacific/Auckland")
with_tz(meeting, "America/Chicago")

```

So the meetings occurred at 4:00 Hadley’s time (and the day before no less). Of course, this was the same actual moment of time as 9:00 in New Zealand. It just appears to be a different day due to the curvature of the Earth.

What if Hadley made a mistake and signed on at 9:00 his time? What time would it then be my time?


```{r}
mistake <- force_tz(meeting, "America/Chicago")
with_tz(mistake, "Pacific/Auckland")
#> [1] "2011-07-02 02:00:00 NZST"
His call would arrive at 2:00 am my time! Luckily he never did that.
```

4. Time Intervals
You can save an interval of time as an Interval class object with lubridate. This is quite useful! For example, my stay in Auckland lasted from June 4, 2011 to August 10, 2011 (which we’ve already saved as arrive and leave). We can create this interval in one of two ways:


```{r}
auckland <- interval(arrive, leave) 
auckland
auckland <- arrive %--% leave
auckland
```

My mentor at the University of Auckland, Chris, traveled to various conferences that year including the Joint Statistical Meetings (JSM). This took him out of the country from July 20 until the end of August.


```{r}
jsm <- interval(ymd(20110720, tz = "Pacific/Auckland"), ymd(20110831, tz = "Pacific/Auckland"))
jsm
```

Will my visit overlap with and his travels? Yes.

```{r}
int_overlaps(jsm, auckland)
```
Then I better make hay while the sun shines! For what part of my visit will Chris be there?

```{r}
setdiff(auckland, jsm)
```
Other functions that work with intervals include int_start, int_end, int_flip, int_shift, int_aligns, union, intersect, setdiff, and %within%.

5. Arithmetic with date times
Intervals are specific time spans (because they are tied to specific dates), but lubridate also supplies two general time span classes: Durations and Periods. Helper functions for creating periods are named after the units of time (plural). Helper functions for creating durations follow the same format but begin with a “d” (for duration) or, if you prefer, and “e” (for exact).

```{r}
minutes(2) ## period
dminutes(2) ## duration
```

Why two classes? Because the timeline is not as reliable as the number line. The Duration class will always supply mathematically precise results. A duration year will always equal 365 days. Periods, on the other hand, fluctuate the same way the timeline does to give intuitive results. This makes them useful for modeling clock times. For example, durations will be honest in the face of a leap year, but periods may return what you want:

```{r}
leap_year(2011) ## regular year
ymd(20110101) + dyears(1)
ymd(20110101) + years(1)
```

```{r}
leap_year(2012) ## leap year
ymd(20120101) + dyears(1)
ymd(20120101) + years(1)
```

You can use periods and durations to do basic arithmetic with date times. For example, if I wanted to set up a reoccuring weekly skype meeting with Hadley, it would occur on:

```{r}
meetings <- meeting + weeks(0:5)
```
Hadley travelled to conferences at the same time as Chris. Which of these meetings would be affected? The last two.


```{r}
meetings %within% jsm
```

How long was my stay in Auckland?

```{r}
auckland / ddays(1)
auckland / ddays(2)
auckland / dminutes(1)
```

And so on. Alternatively, we can do modulo and integer division. Sometimes this is more sensible than division - it is not obvious how to express a remainder as a fraction of a month because the length of a month constantly changes.

```{r}
auckland %/% months(1)
auckland %% months(1)
```
Modulo with an timespan returns the remainder as a new (smaller) interval. You can turn this or any interval into a generalized time span with as.period.

```{r}
as.period(auckland %% months(1))
as.period(auckland)
```

If anyone drove a time machine, they would crash
The length of months and years change so often that doing arithmetic with them can be unintuitive. Consider a simple operation, January 31st + one month. Should the answer be

February 31st (which doesn’t exist)
March 4th (31 days after January 31), or
February 28th (assuming its not a leap year)
A basic property of arithmetic is that a + b - b = a. Only solution 1 obeys this property, but it is an invalid date. I’ve tried to make lubridate as consistent as possible by invoking the following rule if adding or subtracting a month or a year creates an invalid date, lubridate will return an NA. This is new with version 1.3.0, so if you’re an old hand with lubridate be sure to remember this!

If you thought solution 2 or 3 was more useful, no problem. You can still get those results with clever arithmetic, or by using the special %m+% and %m-% operators. %m+% and %m-% automatically roll dates back to the last day of the month, should that be necessary.

```{r}
jan31 <- ymd("2013-01-31")
jan31 + months(0:11)
floor_date(jan31, "month") + months(0:11) + days(31)
jan31 %m+% months(0:11)
```

Notice that this will only affect arithmetic with months (and arithmetic with years if your start date it Feb 29).

6. Vectorization
The code in lubridate is vectorized and ready to be used in both interactive settings and within functions. As an example, here is a function for advancing a date to the last day of the month

```{r}
last_day <- function(date) {
  ceiling_date(date, "month") - days(1)
}
```

### Data Frame

A data frame is a very important data type in R. It's pretty much the *de facto*
data structure for most tabular data and what we use for statistics.

A data frame is a *special type of list* where every element of the list has same length (i.e. data frame is a "rectangular" list).

Data frames can have additional attributes such as `rownames()`, which can be
useful for annotating data, like `subject_id` or `sample_id`. But most of the
time they are not used.

Some additional information on data frames:

* Usually created by `read.csv()` and `read.table()`, i.e. when importing the data into R.
* Assuming all columns in a data frame are of same type, data frame can be converted to a matrix with data.matrix() (preferred) or as.matrix(). Otherwise type coercion will be enforced and the results may not always be what you expect.
* Can also create a new data frame with `data.frame()` function.
* Find the number of rows and columns with `nrow(dat)` and `ncol(dat)`, respectively.
* Rownames are often automatically generated and look like 1, 2, ..., n. Consistency in numbering of rownames may not be honored when rows are reshuffled or subset.

### Creating Data Frames by Hand

To create data frames by hand:

```{r}
dat <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)
dat
```

> ## Useful Data Frame Functions
>
> * `head()` - shows first 6 rows
> * `tail()` - shows last 6 rows
> * `dim()` - returns the dimensions of data frame (i.e. number of rows and number of columns)
> * `nrow()` - number of rows
> * `ncol()` - number of columns
> * `str()` - structure of data frame - name, type and preview of data in each column
> * `names()` - shows the `names` attribute for a data frame, which gives the column names.
> * `sapply(dataframe, class)` - shows the class of each column in the data frame
{: .callout}

See that it is actually a special list:

```{r}
is.list(dat)
class(dat)
```

Because data frames are rectangular, elements of data frame can be referenced by specifying the row and the column index in single square brackets (similar to matrix).

```{r}
dat[1, 3]
```

As data frames are also lists, it is possible to refer to columns (which are elements of such list) using the list notation, i.e. either double square brackets or a `$`.

```{r}
dat[["y"]]
dat$y
```

The following table summarizes the one-dimensional and two-dimensional data structures in R in relation to diversity of data types they can contain.

| Dimensions | Homogenous | Heterogeneous |
| ------- | ---- | ---- |
| 1-D | atomic vector | list |
| 2-D | matrix | data frame |

> Lists can contain elements that are themselves muti-dimensional (e.g. a lists can contain data frames or another type of objects). Lists can also contain elements of any length, therefore list do not necessarily have to be "rectangular". However in order for the list to qualify as a data frame, the lenghth of each element has to be the same. 
{: .callout}


> ## Column Types in Data Frames
>
> Knowing that data frames are lists, can columns be of different type?
>
> What type of structure do you expect to see when you explore the structure of the `iris` data frame? Hint: Use `str()`.
>
> ~~~
> # The Sepal.Length, Sepal.Width, Petal.Length and Petal.Width columns are all
> # numeric types, while Species is a Factor.
> # Lists can have elements of different types.
> # Since a Data Frame is just a special type of list, it can have columns of
> # differing type (although, remember that type must be consistent within each column!).
> str(iris)
> ~~~
> {: .language-r}
{: .challenge}

{% include links.md %}
